---
title: "tapdance_eda"
author: "Natalie Ghidali"
date: "4/15/2021"
output: 
  html_document:
      number_sections: true
      toc: true
      theme: united
      code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# loading packages and set seed
```{r, message = FALSE}
library(tidyverse)
library(tidymodels)
library(skimr)
library(janitor)

set.seed(42)
```

# Initial overview of data

### Source

We will be working with data on the amount of aid given through the Tuition Assistance Program (TAP) for New York universities from the years 2000-2019. The TAP is New York's largest financial aid grant program, assisting eligible New York residents pursue higher education. The dataset includes the dollar amounts of aid given to each recipient, as well as information regarding each recipients' family income, age group, university sector (e.g., public or private), level of study, and more. The dataset was provided by the New York State Higher Education Services Corporation, and it was downloaded from a New York state government data catalog. The dataset has about 288,000 rows, but we intend on focusing only on the years 2014-2019, which will reduce its size. 

The original dataset can be accessed through the following link: https://data.ny.gov/Education/Tuition-Assistance-Program-TAP-Recipients-Dollars-/2t78-bs45

### Load and clean dataset
```{r, message = FALSE}
tap_data <- read_csv("data/tap_dat.csv") %>% 
  clean_names() %>%
  mutate(
    level = factor(level),
    tap_level_of_study = factor(tap_level_of_study),
    sector_type = factor(sector_type),
    tap_sector_group = factor(tap_sector_group),
    recipient_age_group = ordered(recipient_age_group, 
                                  levels = c("under age 22",  "age 22 - 25", "age 26 - 35",  "age 36 - 50",  "over age 50")),
    tap_financial_status = factor(tap_financial_status),
    tap_award_schedule = factor(tap_award_schedule),
    tap_degree_or_non_degree = factor(tap_degree_or_non_degree),
    tap_schedule_letter = factor(tap_schedule_letter),
    income = as.numeric(sub(",","",str_extract(income_by_1_000_range, "\\d+,*\\d*")))
  ) %>%
  select(-income_by_1_000_range, -income_by_5_000_range, -income_by_10_000_range) %>% 
  filter(academic_year >= 2014)
```


### Skim

The resulting dataset contains 56,846 observations and 14 features. Of the 14 variables, 9 are factors and 5 are numeric. The variable `tap_recipient_ft_es`, which refers to the number of TAP recipients. There are no instances of missingness in the dataset.

```{r}
# skim full dataset
skim(tap_data)
```

### Skim outcome variable

```{r}
tap_data %>% ggplot(aes(sector_type, fill = sector_type)) +
  geom_bar() +
  scale_fill_discrete(
    guide = FALSE
  ) +
  labs(
    title = "Institution Type: Private vs. Public"
  ) +
  theme_bw()
```

```{r}
skim(tap_data$sector_type)
```

# Thorough univariate investigation of response and predictor variables

### Academic Year
```{r}
ggplot(tap_data, aes(academic_year)) +
  geom_histogram(bins = 20)
```

Academic year is right skewed, there do not appear to be significant outliers.

### Level
```{r}
ggplot(tap_data, aes(level)) +
  geom_bar()
```

We can see there is a severe class imbalance for the level variable with far more undergraduate students than graduate students.

### TAP Level of Study
```{r}
ggplot(tap_data, aes(tap_level_of_study)) +
  geom_bar()
```

We can see there is a severe class imbalance with far more students in the lower levels of the education system, which makes sense.

### Sector Type
```{r}
ggplot(tap_data, aes(sector_type)) +
  geom_bar()
```

Sector type is fairly balanced, with a similar number of private and public institutions.

### TAP Sector Group
```{r}
ggplot(tap_data, aes(tap_sector_group)) +
  geom_bar() +
  coord_flip()
```

Tap sector group is imbalanced.

### Recipient Age Group
```{r}
ggplot(tap_data, aes(recipient_age_group)) +
  geom_bar()
```

Recipient age group has a fairly normal distribution, with perhapse a slight right skew.

### TAP Financial Status
```{r}
ggplot(tap_data, aes(tap_financial_status)) +
  geom_bar()
```

TAP financial status is imbalanced, with more financially independent cases than financially dependent.

### TAP Award Schedule
```{r}
ggplot(tap_data, aes(tap_award_schedule)) +
  geom_bar()
```

TAP award schedule is imbalanced, there are far more dependently scheduled the independently scheduled or married with no dependents.

### TAP Degree or NonDegree
```{r}
ggplot(tap_data, aes(tap_degree_or_non_degree)) +
  geom_bar()
```

Tap degre or non degree is very imbalanced, with far more counts of degree then non degree.

### TAP Schedule Letter
```{r}
ggplot(tap_data, aes(tap_schedule_letter)) +
  geom_bar()
```

TAP schedule letter is very imbalanced. There are many Es in comparison to a relatively low count for other classes.

### Income
```{r}
ggplot(tap_data, aes(income)) +
  geom_histogram(bins = 20)
```

There are many more students at the lower end of the income spectrum vs at the higher end. Incomes do not exceed $80,000.

### TAP Recipient Headcount
```{r}
ggplot(tap_data, aes(tap_recipient_headcount)) +
  geom_histogram(bins = 50) +
  xlim(c(0,50))
```

The TAP recipient headcount is heavily right skewed with a very long tail.


### TAP Recipient FTEs
```{r}
ggplot(tap_data, aes(tap_recipient_ft_es)) +
  geom_histogram() +
  xlim(0,50)
```

The TAP recipient ftes is heavily right skewed with a very long tail.

### TAP Recipient Dollars
```{r}
ggplot(tap_data, aes(tap_recipient_dollars)) +
  geom_histogram() +
  xlim(0,200000)
```

The TAP recipient dollars is heavily right skewed with a very long tail.

# Interesting/important relationships between and among response variable(s) and predictor variables
```{r}

```

# General variable exploration
```{r}

```

## Relationships between response variables and predictor variables

We first wanted to see the distribution of the total amount of aid given to students attending private or public schools. 

```{r}
# Relationship between sector type and amount of aid awarded
ggplot(tap_data, aes(x = sector_type, y = tap_recipient_dollars)) +
  geom_bar(stat = "identity") +
  labs(x = "Sector Type", y = "Total Aid Given (USD)", title = "Total TAP Aid Given by Sector Type") +
  theme_minimal()
```

There was more total aid allotted to students attending public schools than private schools

```{r}
# Income Level vs Sector Type
ggplot(tap_data, aes(x = sector_type, y = income)) +
  geom_boxplot() +
  labs(x = "Sector Type", y = "Income in USD", title = "Average Income Level Among Students by Sector Type") +
  theme_minimal()
```

There is s higher average income among students attending public schools, which raises questions about whether more public school students are financially dependent, then why do public school students receive more aid overall. However, the average income is only slightly higher among public school students than private school students, so we don't know to what exent this difference is important. 

```{r}
# Financial Status and income

ggplot(tap_data, aes(x = tap_financial_status, y = income)) +
  geom_boxplot() +
  labs(x = "Financial Status", y = "Income (USD)", title = "Average Income of TAP Recipients by Financial Status") +
  theme_minimal()
```

Dependents have an average higher income than students who are financially independent

```{r}
# Income vs TAP Money Given
ggplot(tap_data, aes(x = income, y = tap_recipient_dollars)) +
  geom_line() +
  labs(x = "Income (USD)", y = "Amount of TAP Aid Given (USD)", title = "Amount of Aid Given By Income") +
  theme_minimal()
```

The amount of aid given to students decreases as their level of income increases. 

